Compilation and Backend-Independent Vectorization for Multi-Party Computation

### 2022-04-27 overleaf/paper_SIMD.tex

@@

\subsection{Our contribution} In this paper, we develop a compiler framework that takes a Python-like routine and produces MOTION code: we describe (a) the IMP-MPC language, its syntax and semantic restrictions, (b) translation from IMP-MPC into MPC Source, (c) a specific circuit-independent optimization: novel SIMD-vectorization on MPC Source, and (d) translation from MPC Source into MOTION code. \ana{I think we have to add more on MOTION, specifically mixing and amortization.}

Our second contribution is an analytical model for cost estimation of amortized schedules. Originally, we hoped that optimal scheduling (under our model) was tractable, as the problem
appeared simpler than the classical scheduling problem. Unfortunately, we show that optimal scheduling is NP-hard via a reduction to the Shortest Common Supersequence (SCS) problem. 
Cost estimation is important as it drives scheduling and protocol mixing optimizations~\cite{Ishaq:2019, Fang:2022}.

@@

@@

As a running example, consider Biometric matching, a standard MPC benchmark.
Array \texttt{C} is the feature vector of \texttt{D} features that we wish to match and array \texttt{S} 
is the database of \texttt{N} vectors of size \texttt{D} that we match against.
An intuitive implementation is as follows:

{\small
\begin{verbatim}
def biometric(C: shared[list[int]], D: int, 
              S: shared[list[int]], N: int) -> 
                      tuple[shared[int], shared[int]]:
    min_sum = 10000
    min_index = 0  
    for i in range(N): #loop over database
        sum = 0
        for j in range(D): #loop over features
            d = S[i * D + j] - C[j] #i.e., d = S[i,j] - C[j]
            p = d * d
            sum = sum + p
        if sum < min_sum:
            min_sum = sum
            min_index = i
    return (min_sum, min_index)
\end{verbatim}
}

Our compiler takes (essentially) standard IMP syntax.
The programmer can write intuitive iterative programs as the one above.
They annotate certain inputs and outputs as \emph{shared}.
Here the code iterates over the entries in the database and computes
the sum of squares of the differences of individual features. The program returns the 
index \texttt{i} of the vector that gives the best match plus the corresponding sum of squares.

@@

The compiler generates an IR, MPC source:
{\small
\begin{verbatim}
1. min_sum!1 = 10000
2. min_index!1 = 0
3. for i in range(0, N!0):
4.    min_sum!2 = PHI(min_sum!1, min_sum!4)
5.    min_index!2 = PHI(min_index!1, min_index!4)
6.    sum!2 = 0
7.    for j in range(0, D!0):
8.      sum!3 = PHI(sum!2, sum!4)
9.      d!3 = (S!0[((i * D!0) + j)] - C!0[j]) // MPC
10.     p!3 = (d!3 * d!3) // MPC
11.     sum!4 = (sum!3 + p!3) // MPC
12.  !1!2 = (sum!3 < min_sum!2) // MPC
13.  min_sum!3 = sum!3
14.  min_index!3 = i
15.  min_sum!4 = MUX(!1!2, min_sum!3, min_sum!2) // MPC
16.  min_index!4 = MUX(!1!2, min_index!3, min_index!2) // MPC
17.!2!1 = (min_sum!2, min_index!2)   
\end{verbatim}
}

The compiler linearizes the source turning conditionals into MUX statements. The PHI nodes
are remnants of the SSA IR; the compiler generates code that picks the correct value 
when producing MOTION output; the MOTION framework in turn linearizes 
loops when it generates the circuit.

We turn to our analytical model to compute the cost of this program. Assuming fixed 
cost $\beta$ for a local MPC operation (essentially just ADD) and cost $\alpha$ for a remote
MPC operation (e.g., MUX, CMP, and remaining operations), the cost of the iterative schedule 
will be $N*D*(2*\alpha+\beta) + N*3*\alpha$.

A key contribution is the vectorizing transformation. We can compute all $N*D$ 
subtraction operations (line 9) in a single SIMD instruction; similarly we can compute 
all multiplication operations (line 10) in a single SIMD instruction. And while we cannot
vectorize computation of the N individual sums, we can compute the N sums in parallel. 
Our compiler \emph{automatically detects these opportunities and transforms the program}. 
It is standard that MPC researchers write vectorized versions of the Biometric
program by hand; we are the first (to the best of our knowledge) to automatically 
transform an intuitive, iterative MPC program into an unintuitive vectorized one.

@@

Our compiler produces the following vectorized program. (Note that this is
still higher-level IR, Vectorized MPC Source. Our compiler turns this code into MOTION variables, 
loops and SIMD primitives, which MOTION then uses to generate the circuit.)

{\small
\begin{verbatim}
min_sum!1 = 10000
min_index!1 = 0
// S!0^ is same as S!0. C!0^ replicates C!0 N-times: 
S!0^ = raise_dim(S!0, ((i * D!0) + j), (i:N!0,j:D!0))
C!0^ = raise_dim(C!0, j, (i:N!0,j:D!0))

@@

for i in range(0, N!0):
  min_index!2 = PHI(min_index!1, min_index!4)  
  min_index!4 = MUX(!1!2[i], min_index!3[i], min_index!2)
!2!1 = (min_sum!2, min_index!2)   
\end{verbatim}
}

In MPC compilers a vectorized operation that computing M operations "at once" costs 
essentially the same ($\alpha$ or $\beta$) as an individual operation. We elaborate 
on these in the following section. Thus, the vectorized program costs 
$2*\alpha$ + $D * \beta + N*3*\alpha$. The first term in the sum corresponds to the vectorized
subtraction and multiplication, the second term corresponds to the for loop on $j$ and the third 
one corresponds to the remaining for loops on $i$. Clearly, $2*\alpha + D * \beta + N*3*\alpha << N*D*(2*\alpha+\beta) + N*3*\alpha$.
Our experimental results illustrate this as well. \ana{Add numbers.}

@@

>>>>

\ana{We need an abstract.}

@@

\subsection{Our Contribution} In this paper, we develop a compiler framework that takes a Python-like routine and produces MOTION code: we describe (a) the IMP-MPC language, its syntax and semantic restrictions, (b) translation from IMP-MPC into MPC Source, (c) a specific circuit-independent optimization: novel SIMD-vectorization on MPC Source, and (d) translation from MPC Source into MOTION code. %\ana{I think we have to add more on MOTION, specifically mixing and amortization.}

We focus on the MOTION framework as our back-end for several reasons. First, it is the state of the art in terms of performance~\cite{Braun:2022}. Second, it provides an API over efficient implementation for a wide variety of cryptographic operations in three different protocols ---  Arithmetic GMW, Boolean GMW, and BMR --- which allows for protocol mixing~\cite{Ishaq:2019, Fang:2022}, a known backend-independent optimization. Third, MOTION provides API for SIMD-level operations, which amortize cost and lead to significant improvement in memory footprint and 
throughput~\cite{NDSS:DemSchZoh15, CCS:ABFKLO18, Braun:2022}. It also enables MPC Source-level optimizations such SIMD vectorization.

Our second contribution is an analytical model for cost estimation of amortized schedules. Originally, we hoped that optimal scheduling (under our model) was tractable, as the problem
appeared simpler than the classical scheduling problem. Unfortunately, we show that optimal scheduling is NP-hard via a reduction to the Shortest Common Supersequence (SCS) problem. 
Cost estimation is important as it drives not only vectorization but other optimizations such as protocol mixing optimizations and scheduling as well~\cite{Ishaq:2019, Fang:2022}.

@@

\begin{table*}
\begin{tabular}{ccc}
\begin{minipage}{0.3\textwidth}
{\small
\begin{pythonn}
def biometric(C: shared[list[int]], D: int, 
      S: shared[list[int]], N: int) -> 
      shared[tuple[int,int]]:
   min_sum = MAX_INT
   min_idx = 0  
   for i in range(N): 
      sum = 0
      for j in range(D): 
         # d = S[i,j] - C[j]
         d = S[i * D + j] - C[j] 
         p = d * d
         sum = sum + p
      if sum < min_sum:
         min_sum = sum
         min_idx = i
   return (min_sum, min_idx)
\end{pythonn}
}
\end{minipage}

&

\begin{minipage}{0.33\textwidth}
{\tiny
\begin{pythonn}
min_sum!1 = MAX_INT
min_idx!1 = 0
for i in range(0, N):
   min_sum!2 = PHI(min_sum!1, min_sum!4)
   min_idx!2 = PHI(min_idx!1, min_idx!4)
   sum!2 = 0
   for j in range(0, D):
      sum!3 = PHI(sum!2, sum!4)
      d = SUB(S[((i * D) + j)],C[j]) 
      p = MUL(d,d) 
      sum!4 = ADD(sum!3,p) 
   t = CMP(sum!3,min_sum!2) 
   min_sum!3 = sum!3
   min_idx!3 = i
   min_sum!4 = MUX(t, min_sum!3, min_sum!2) 
   min_idx!4 = MUX(t, min_idx!3, min_idx!2) 
return (min_sum!2, min_idx!2)   
\end{pythonn}
}
\end{minipage}

&

\begin{minipage}{0.33\textwidth}
{\small
\begin{pythonn}
min_sum!1 = MAX_INT
min_idx!1 = 0
# S^ is same as S. C^ replicates C N times: 
S^ = lift(S, ((i * D) + j), (i:N,j:D)) # S^[i,j] = S[i,j]
C^ = lift(C, j, (i:N,j:D)) # C^[i,j] = C[j]

sum!2[I] = [0,..,0]
# computes _all_  "at once"
d[I,J] = SUB_SIMD(S^[I,J],C^[I,J])
p[I,J] = MUL_SIMD(d[I,J],d[I,J])

for j in range(0, D):
   # sum!2[I], sum!3[I], sum!4[I] are size-N vectors 
   # computes N intermediate sums "at once"
   sum!3[I] = PHI(sum!2[I], sum!4[I])       
   sum!4[I] = ADD_SIMD(sum!3[I],p[I,j])

min_idx!3[I] = [0,1,...N-1]   
for i in range(0, N):
   min_sum!2 = PHI(min_sum!1, min_sum!4) 
   t[i] = CMP(sum!3[i],min_sum!2)
   min_sum!4 = MUX(t[i], sum!3[i], min_sum!2)  
for i in range(0, N):
   min_idx!2 = PHI(min_idx!1, min_idx!4)    
   min_idx!4 = MUX(t[i], min_idx!3[i], min_idx!2)
return (min_sum!2, min_idx!2)   
\end{pythonn}
}
\end{minipage}

\\

(a) IMP Source & (b) MPC Source & (c) Optimized MPC Source
\end{tabular}
\caption{Biometric Matching: From IMP Source to MPC Source to Optimized MPC Source. }
\label{tab:source_and_MPC_source_and_optimized_MPC_source}
\end{table*}

@@

As a running example, consider Biometric matching, a standard MPC benchmark.
An intuitive (and naive) implementation is as shown in Listing~\ref{tab:source_and_MPC_source_and_optimized_MPC_source}(a).
Array \python{C} is the feature vector of \texttt{D} features that we wish to match and \python{S} 
is the database of \python{N} size-\python{D} vectors that we match against.

\begin{comment}
{\small
\begin{verbatim}
def biometric(C: shared[list[int]], D: int, 
              S: shared[list[int]], N: int) -> 
                      tuple[shared[int], shared[int]]:
    min_sum = 10000
    min_index = 0  
    for i in range(N): #loop over database
        sum = 0
        for j in range(D): #loop over features
            d = S[i * D + j] - C[j] #i.e., d = S[i,j] - C[j]
            p = d * d
            sum = sum + p
        if sum < min_sum:
            min_sum = sum
            min_index = i
    return (min_sum, min_index)
\end{verbatim}
}
\end{comment}

Our compiler takes essentially standard IMP~\ana{Need citation here.} 
syntax and imposes certain semantic restrictions. (We detail the restrictions in the following sections.)
The programmer writes an iterative program and annotate certain inputs 
and outputs as \emph{shared}. In the example arrays \python{C} and \python{S}
are shared, meaning that they store shares, however, the array sizes \python{D} and
\python{N} respectively are plaintext. The code iterates over the entries in the 
database and computes the Euclidean distance (its square actually) of the current 
entry \python{S[i]} and \python{C}. The program returns the index of the vector that gives 
the best match plus the corresponding sum of squares.

@@

Our compiler generates an intermediate representation, MPC Source. MPC Source 
for Biometric Matching is shown in 
Listing~\ref{tab:source_and_MPC_source_and_optimized_MPC_source}(b). First, MPC Source is an SSA form. 
Second, it is linear. The conditional in lines 13-14 in IMP Source turns into the linear code in lines 12-16 in MPC Source. 
The test turns into a CMP operation (\python{t = CMP(sum!3,min\_sum!2)}), followed by the 
true-branch sequence, followed by the MUX operations. The first MUX operation selects the value
of \python{min\_sum}: if \python{t} is true, then \python{min\_sum} gets the value of the second multiplexer
argument,  \python{min\_sum!3}, otherwise it takes the value of the third argument, \python{min\_sum!2}.  
Third, MPC Source is a special form of SSA. The SSA $\phi$-nodes at the if-then-else (lines 13-15) turn into 
MUX operations, while the $\phi$-nodes at for loops turn into what we call pseudo PHI nodes with a straightforward semantics.

\begin{comment}
{\small
\begin{pythonn}
1. min_sum!1 = 10000
2. min_index!1 = 0
3. for i in range(0, N!0):
4.    min_sum!2 = PHI(min_sum!1, min_sum!4)
5.    min_index!2 = PHI(min_index!1, min_index!4)
6.    sum!2 = 0
7.    for j in range(0, D!0):
8.      sum!3 = PHI(sum!2, sum!4)
9.      d!3 = (S!0[((i * D!0) + j)] - C!0[j]) // MPC
10.     p!3 = (d!3 * d!3) // MPC
11.     sum!4 = (sum!3 + p!3) // MPC
12.  !1!2 = (sum!3 < min_sum!2) // MPC
13.  min_sum!3 = sum!3
14.  min_index!3 = i
15.  min_sum!4 = MUX(!1!2, min_sum!3, min_sum!2) // MPC
16.  min_index!4 = MUX(!1!2, min_index!3, min_index!2) // MPC
17.!2!1 = (min_sum!2, min_index!2)   
\end{pythonn}
}
\end{comment}

%The compiler linearizes the source turning conditionals into MUX statements. The PHI nodes
%are remnants of the SSA IR; the compiler generates code that picks the correct value 
%when producing MOTION output; the MOTION framework in turn linearizes 
%loops when it generates the circuit.

We turn to our analytical model to compute the \emph{cost} of the program. Assuming fixed 
cost $\beta$ for a local MPC operation (essentially just ADD) and cost $\alpha$ for a remote
MPC operation (e.g., MUX, CMP, and remaining operations), the cost of the iterative schedule 
will be $ND(2\alpha+\beta) + N(3\alpha$).

A key contribution is the vectorizing transformation. We can compute all $N*D$ 
subtraction operations (line 9) in a single SIMD instruction; similarly we can compute 
all multiplication operations (line 10) in a single SIMD instruction. And while we cannot
vectorize computation of the N individual sums, we can compute the N sums in parallel. 
Our compiler \emph{automatically detects these opportunities and transforms the program}. 
It is standard that MPC researchers write vectorized versions of the Biometric
program by hand; we are the first (to the best of our knowledge) to automatically 
transform a naive, iterative MPC program into an unintuitive vectorized one.

@@

Our compiler produces the vectorized program shown in Listing~\ref{tab:source_and_MPC_source_and_optimized_MPC_source}(c). 
Note that this is still our intermediate representation, Optimized MPC Source. Subsequently,
the compiler turns this code into MOTION variables, loops and SIMD primitives, which MOTION then 
uses to generate the circuit.

The compiler determines that SUB and MUL in ``naive'' MPC Source (lines 9 and 10 in (b))
can be fully vectorized into the SIMD SUB and MUL in optimized MPC Source (lines 9 and 10 in (c)). 
Notation \python{p[I,J]} denotes a 2-dimensional array with fully vectorized dimensions.
The computation of sum (line 11 in (b))
is sequential across the $j$-dimension, but it is parallel across the $i$-dimension. 
The loop in lines 12-16 in (c) illustrates; here \python{p[I,j]} refers to the $j$-th column in \python{p}. 
Unfortunately, CMP and MUX remain sequential.

\begin{comment}
{\small
\begin{verbatim}
min_sum!1 = 10000
min_index!1 = 0
// S!0^ is same as S!0. C!0^ replicates C!0 N-times: 
S!0^ = raise_dim(S!0, ((i * D!0) + j), (i:N!0,j:D!0))
C!0^ = raise_dim(C!0, j, (i:N!0,j:D!0))

@@

for i in range(0, N!0):
  min_index!2 = PHI(min_index!1, min_index!4)  
  min_index!4 = MUX(!1!2[i], min_index!3[i], min_index!2)
!2!1 = (min_sum!2, min_index!2)   
\end{verbatim}
}
\end{comment}

In MPC backends, executing $n$ operations ``at once'' in a single SIMD operation costs a lot less than executing those $n$ operations one by one.
This is particularly important when there is communication (i.e., in remote), since many 1-bit values are sent at once rather than sequentially. 
%In MPC compilers a vectorized operation that computing M operations "at once" costs 
%essentially the same ($\alpha$ or $\beta$) as an individual operation. 
We elaborate on the cost model in the following section but for now consider that 
each operation has a \emph{fixed} portion (does benefits from amortization) and 
a \emph{variable} portion (does not benefit from amortization): $\alpha = \alpha_\mathit{fix} + \alpha_\mathit{var}$. 
This gives rise to the following formula for amortized cost: $f(n) = \alpha_\mathit{fix} + n\alpha_\mathit{var}$, 
as opposed to unamortized cost $g(n) = n\alpha_\mathit{fix} + n\alpha_\mathit{var}$. (We extend the same reasoning to 
$\beta$-instructions.)

Thus, the fixed cost of the vectorized program amounts to
$2\alpha_\mathit{fix}$ + $D \beta_\mathit{fix} + N(3\alpha_\mathit{fix})$. (The variable cost is the same in both the vectorized and un-vectorized programs.) 
The first term in the sum corresponds to the vectorized
subtraction and multiplication, the second term corresponds to the FOR loop on $j$ and the third 
one corresponds to the remaining FOR loops on $i$. Clearly, $2\alpha_\mathit{fix} + D\beta_\mathit{fix} + N(3\alpha_\mathit{fix}) << ND(2\alpha_\mathit{fix}+\beta_\mathit{fix}) + N3\alpha_\mathit{fix}$.
Empirically, we observe 10x to 50x improvement over un-vectorized Biometric Matching in circuit generation time and setup time and 5x to 30x improvement in online time.
Additionally, the un-vectorized version runs out of memory for N=256, while the vectorized one continues to run up to standard maximal input size N=4096.
Our experimental results illustrate this as well. \ana{Check numbers.}

@@

\subsection{Syntax}
